A free/open-source Kazakh-Tatar machine translation system

### 2013-04-22 paper/paper.tex
cooperate with spectie
updates and a coupe of otodos

@@

\begin{document}
\spacing{1}
\maketitle
\begin{abstract}
This paper presents a bidirectional machine translation system between Kazakh and Tatar.
\end{abstract}
%FIXME
%\maketitleabstract

@@

Within the Apertium project, work on MT systems between Turkic languages has been started (Turkish-Kyrgyz, Azeri-Turkish), but the Kazakh-Tatar system described by the present study is the closest to production-ready of them.  Among these systems is a prototype Tatar-Bashkir machine translation system which was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.

Besides these systems, several previous works on making machine translation systems between Turkic languages exist, although to our knowledge none are publically available.
Some MT systems have been reported that translate between Turkish and other Turkic languages, including Turkish-Crimean Tatar \citep{altintas01},
Turkish-Azerbaijani \citep{hamzaoglu93}, Turkish-Tatar \citep{suleymanov08}, and
Turkish-Turkmen \citep{tantug07}, though none of these have been released to a public audience.

@@

Both Tatar and Kazakh belong to the Kypchak (or Northwestern) group of Turkic
languages.  The spoken and written languages share some level of mutual intelligibility to naïve native speakers, though this is somewhat limited, and is obscured by different orthographical conventions and some opaque correspondences.

@@

While Tatar and Kazakh both have a velar/uvular obstruent distinction (e.g., /k/ vs.\ /q/) that interacts with adjacent vowels, the Tatar orthography only has one series of letters (e.g., ‹к›), relying on adjacent vowels (and employing ``hard'' and ``soft signs'' when these fail) to differentiate the two, and Kazakh has two series of obstruents (e.g., ‹к› and ‹қ›).

Kazakh does not orthographically distinguish high unrounded vowels (/{\qipa ɘ}/ ‹і› and /ə/ ‹ы›) before glides (/w/ ‹у› and /j/ ‹й›) by writing the combination with one letter; i.e., /{\qipa ɘ}j/ and /əj/ are both written ‹и›, while /{\qipa ɘ}w/ and /əw/ are both written ‹у›.  The quality of these vowels is necessary to know in order to predict the quality of following harmonising vowels. %, so this convention presents accute challenges to producing an accurate morphological transducer.
Additionally, Tatar and Kazakh both use ``yoticised'' vowels---i.e., when ‹о›, ‹у›, or ‹а› (along with ‹ә› in Tatar) follow /j/, a single character is used to represent both: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›, though ‹е› is also the non-word-initial variant of ‹э›.}  %Combined with the issues regarding ‹у› in Kazakh,

All of these orthographical conventions present accute challenges to designing accurate morphological transducers for the languages.

\subsection{Morphological differences}
There are a number of examples where the morphologies of Kazakh and Tatar are rather different, including morphemes in one language that don't exist in the other, entirely different uses of the same morpheme combinations, and morphotactic differences (i.e., allowable ordering and placement of morphemes).

An example of a morpheme that doesn't exist in one of the two languages is Kazakh -\texttt{\{E\}т\{I\}н}, which is used to form non-past verbal adjectives and verbal nouns.  The semantically equivalent structure in Tatar is -\texttt{\{E\} торган}, which historically corresponds to the source of the Kazakh morpheme; however, the use of -\texttt{\{E\} тұрған} in modern Kazakh is different from that of -\texttt{\{E\}т\{I\}н}.

@@

This systematic difference would seem to be a minor issue, since, as is typical in pro-drop languages, pronouns are only used for emphasis and clarification.  However, this difference between Tatar and Kazakh in the second person system runs much deeper than just the pronoun system.  Since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (even when there is no overt pronoun, in either situation), the Kazakh and Tatar systems of agreement suffixes reflect the same pattern; i.e., there are several sets of agreement morphemes which have a one-to-one correspondence with the pronouns in each language, resulting in several systems of suffixes in each language that have the same set of distinctions as in the 2nd person pronoun systems.

@@

Without regard to the semantic alignment of these forms, the morphotactics of the cognate Kazakh distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are different.  Specifically, in both languages, the person agreement takes the form of a person copula suffix, although in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in Tatar this suffix intervenes between the two pieces of the ``compound'' tense morpheme (e.g., баргансың икән ``I guess you went'').
% I think that the meaning of the Tatar "баргансың икән" is also "It turns out you went, It appears that you went,
% They say you went", i.e. it is closer to Kazakh. "I guess you went" sounds like you are supposing that, which is not the case.
% Maybe I just don't know the whole semantics of the english "to guess". If you are sure about it, just leave it as it is.

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:GAlI}
\a
\begingl
\gla Мен сенімен сөйлескелі келдім. //
\glb мен сен-I-мен сөйле-с-GAлI кел-DI-м. //
\glb I you-\gmk{Acc}-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\a
\begingl
\gla Мин синең белән сөйлешергә килдем. //
\glb мин син-Iң белән сөйле-ш-IргA кил-DI-м. //
\glb I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\xe

@@

The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly by hand (i.e., by translating Kazakh stems unrecognized by the morphological analyser into Tatar).  Some toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia; also, some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.) were added to the bilingual dictionary automatically by taking the intersection of Russian and Kazakh wordlists.

@@

%FIXME: JNW needs to copy-edit this, and we should probably provide a concrete example to make it clearer too
The output of the morphological analyser is highly ambigious. Both Kazakh and Tatar have a lot of affixes which join the
verb stem. Some of the grammars label them all gerunds, or participles, assuming that there are no finite forms.

We give to a VerbStem+ParticularAffix surface form maximal number of readings (syntactic roles) it can have, such as [verbal
adjective], [finite form], [substantivized verbal adjective, i.e., if the word it determines was ommitted], whereas many grammars
do not distinguish these and label forms of verbs containing a particular affix just as a certain gerund or participle form.

This type of ambiguity is common between the two languages we deal with, and therefore may be passed from one language
to another and most likely won't lead to translation errors. Disambiguating it, and, prior to that, having that kind of
informaton about the syntactical role of a surface form is crucial for success e.g. of a Turkic to English system.

@@

%\^{}Һауа\tag{<n>}\tag{<nom>\$} \^{}бөгөн\tag{<adv>\$} \^{}бик\tag{<adv>\$} \^{}әйбәт\tag{<adj>\$}\^{},\tag{<cm>\$} \\ 
%~                    & \^{}йылы\tag{<adj>\$} \^{}ғына\tag{<postadv>\$}\^{}.\tag{<sent>\$}\\
\midrule
{\bf Mor. generation} & Ул инде ул тавышны аныграк ишетә башлады. \\
\bottomrule
\end{tabular}
 \caption{Translation process for the phrase \emph{Ол енді ол дыбысты анығырақ ести бастады} \eng{He begins to listen to that sound more carefully}.}
 %\vspace{-1em}
\end{table*}

@@

The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.

@@

As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT system\footnote{Note that the coverage of the vanilla transducers is somewhat higher.} over the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.  These unknown words are principally proper nouns.  For the Wikipedia corpus, the coverage is substantially worse.  This is due to the fact that this corpus is ``dirtier'', as well as to the fact that it has contains many more proper nouns.

To measure the performance of the translator we used the Word Error Rate metiric --- an edit metric based on the
Levenshtein distance \citep{levenshtein/1966}.

We had two small Kazakh corporas alongside with their postedited translations into Tatar to measure the WER.
The first one (2457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
the system, and all the rules (Constraint Grammar, lexical selection and transfer) were based on this corpus.
Hence we can call it the development corpus. It presents an upper bound on the current performance
of the system. Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN

The second corpus --- a collection of only Radio Free Europe / Radio Liberty's
Kazakh-language service articles --- was used solely for evaluation. It had a similar size --- 2862 words.
The WER results for it can be seen in the table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.

@@

@@

To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has a near-to-production-level
coverage, but rather a prototype-level number of rules. Though the impact of this relatively low number of rules on the
quality of the translation (compare the WER results for the corpus 1 and corpus 2) looks very promising and suggests
the a high-quality translation between morphologically rich agglutinative languages is possible.

We plan to continue development on the pair; the coverage of the system is already quite high, and although we intend to
increase it to 95% on the corporas we have, the main work will be improving the quality of the translation by adding
more rules, starting with a Constraint Grammar module.
The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make
transfer systems between all the Turkic language pairs.  Related work is currently ongoing with Chuvash--Turkish and Turkish--Kyrgyz.

@@

The work on this Kazakh-Tatar machine translation system was partially funded by the Google Summer of Code and Google Code-In programs.

>>>>

\newcommand{\todo}[1]{\{\textbf{TODO: #1}\}}

@@

\begin{document}
\spacing{1}
\maketitle
\begin{abstract}
This paper presents a bidirectional machine translation system between Kazakh and Tatar.
\todo{expand abstract}
\end{abstract}
%FIXME
%\maketitleabstract

@@

Within the Apertium project, work on MT systems between Turkic languages has been started (Turkish--Kyrgyz, Azeri--Turkish, Turkish-Bashkir), but the Kazakh--Tatar system described by the present study is the closest to production-ready of them.  Among these systems is a prototype Tatar-Bashkir machine translation system which was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.

Besides these systems, several previous works on making machine translation systems between Turkic languages 
exist, although to our knowledge none are publically available.
Some MT systems have been reported that translate between Turkish and other Turkic languages, 
including Turkish--Crimean Tatar \citep{altintas01},
Turkish--Azerbaijani \citep{hamzaoglu93}, Turkish--Tatar \citep{suleymanov08}, and
Turkish--Turkmen \citep{tantug07}, though none of these have been released to a public audience. The Turkish--Azerbaijani
pair is also available through Google Translate.\footnote{\url{http://translate.google.com}}

@@

Both Tatar and Kazakh belong to the Kypchak (or Northwestern) group of Turkic
languages.  The spoken and written languages share some level of mutual intelligibility to native 
speakers, though this is somewhat limited, and is obscured by different orthographical 
conventions and some opaque correspondences.

@@

While Tatar and Kazakh both have a velar/uvular obstruent distinction (e.g., /k/ vs.\ /q/) that interacts with 
adjacent vowels, the Tatar orthography only has one series of letters (e.g., ‹к›), relying on adjacent 
vowels (and employing ъ `hard sign' and ь `soft sign' when these fail) to differentiate the two, and Kazakh 
has two series of obstruents (e.g., ‹к› and ‹қ›).

Kazakh does not orthographically distinguish high unrounded vowels (/{\qipa ɘ}/ ‹і› and /ə/ ‹ы›) before 
glides (/w/ ‹у› and /j/ ‹й›) by writing the combination with one letter; i.e., /{\qipa ɘ}j/ and /əj/ are both 
written ‹и›, while /{\qipa ɘ}w/ and /əw/ are both written ‹у›.  The quality of these vowels is necessary to know in 
order to predict the quality of following harmonising vowels. %, so this convention presents accute challenges to producing an accurate morphological transducer.
Additionally, Tatar and Kazakh both use `yoticised' vowels---i.e., when ‹о›, ‹у›, or ‹а› (along with ‹ә› in 
Tatar) follow /j/, a single character is used to represent both: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›, though ‹е› is also the non-word-initial variant of ‹э›.}  %Combined with the issues regarding ‹у› in Kazakh,

All of these orthographical conventions present acute challenges to designing accurate morphological transducers for the languages.

\subsection{Morphological differences}
There are a number of examples where the morphologies of Kazakh and Tatar are rather different, including 
morphemes in one language that do not exist in the other, entirely different uses of the same morpheme 
combinations, and morphotactic differences (i.e., allowable ordering and placement of morphemes).

An example of a morpheme that does not exist in one of the two languages is Kazakh -\texttt{\{E\}т\{I\}н}, 
which is used to form non-past verbal adjectives and verbal nouns.  The semantically equivalent structure 
in Tatar is -\texttt{\{E\} торган}, which historically corresponds to the source of the Kazakh morpheme; 
however, the use of -\texttt{\{E\} тұрған} in modern Kazakh is different from that of -\texttt{\{E\}т\{I\}н}.

@@

This systematic difference would seem to be a minor issue, since, as is typical in pro-drop languages, pronouns 
are only used for emphasis and clarification.  However, this difference between Tatar and Kazakh in 
the second-person system runs much deeper than just the pronoun system.  Since all finite verb forms 
morphologically agree in person and number with their subject and all possessed nouns agree in person and 
number with their possessor (even when there is no overt pronoun, in either situation), the Kazakh and Tatar 
systems of agreement suffixes reflect the same pattern; i.e., there are several sets of agreement morphemes 
which have a one-to-one correspondence with the pronouns in each language, resulting in several systems of 
suffixes in each language that have the same set of distinctions as in the 2nd person pronoun systems.

@@

Without regard to the semantic alignment of these forms, the morphotactics of the cognate Kazakh 
distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are 
different.  Specifically, in both languages, the person agreement takes the form of a person copula suffix, although 
in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in 
Tatar this suffix intervenes between the two pieces of the `compound' tense morpheme (e.g., баргансың икән ``I guess you went'').
% I think that the meaning of the Tatar "баргансың икән" is also "It turns out you went, It appears that you went,
% They say you went", i.e. it is closer to Kazakh. "I guess you went" sounds like you are supposing that, which is not the case.
% Maybe I just don't know the whole semantics of the english "to guess". If you are sure about it, just leave it as it is.

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\label{ex:GAlI}
\a
\begingl
\gla Мен сенімен сөйлескелі келдім. //
\glb мен сен-мен сөйле-с-GAлI кел-DI-м. //
\glb I you-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\a
\begingl
\gla Мин синең белән сөйлешергә килдем. //
\glb мин син-Iң белән сөйле-ш-IргA кил-DI-м. //
\glb I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\xe

@@

The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly 
by hand (i.e., by translating Kazakh stems unrecognised by the morphological analyser into Tatar).  Some 
toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia \citep{tyers08}; also, 
some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.) were added to the bilingual 
dictionary automatically by taking the intersection of Russian and Kazakh wordlists.

@@

%FIXME: JNW needs to copy-edit this, and we should probably provide a concrete example to make it clearer too
The output of the morphological analyser is highly ambiguous. Both Kazakh and Tatar have a lot of affixes which join the
verb stem. Some of the grammars label them all gerunds, or participles, assuming that there are no finite forms.

We give to a VerbStem+ParticularAffix surface form maximal number of readings (syntactic roles) it can have, such as [verbal
adjective], [finite form], [substantivised verbal adjective, i.e., if the word it determines was omitted], whereas many grammars
do not distinguish these and label forms of verbs containing a particular affix just as a certain gerund or participle form.

This type of ambiguity is common between the two languages we deal with, and therefore may be passed from one language
to another and most likely will not lead to translation errors. Disambiguating it, and, prior to that, having that kind of
informaton about the syntactical role of a surface form is crucial for success e.g. of a Turkic to English system.

@@

%\^{}Һауа\tag{<n>}\tag{<nom>\$} \^{}бөгөн\tag{<adv>\$} \^{}бик\tag{<adv>\$} \^{}әйбәт\tag{<adj>\$}\^{},\tag{<cm>\$} \\ 
%~                    & \^{}йылы\tag{<adj>\$} \^{}ғына\tag{<postadv>\$}\^{}.\tag{<sent>\$}\\
\midrule
{\bf Mor. generation} & Ул инде ул тавышны аныграк ишетә башлады. \\
\bottomrule
\end{tabular}
 \caption{Translation process for the Kazakh phrase \emph{Ол енді ол дыбысты анығырақ ести бастады} \eng{He begins to listen to that sound more carefully}.}
 %\vspace{-1em}
\end{table*}

@@

%The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.

@@

As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT 
system\footnote{Note that the coverage of the vanilla transducers is somewhat higher.} over 
the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.
For the Wikipedia corpus, the coverage 
is substantially worse.  This is due to the fact that this corpus is ``dirtier'' --- 
containing more orthographical errors, \emph{wiki} code, and repetitions, as well as to the 
fact that it has contains many more proper nouns.

To measure the performance of the translator we used the Word Error Rate metric --- an edit-distance metric based on the
Levenshtein distance \citep{levenshtein/1966}.

We had two small Kazakh corpora along with their postedited translations into Tatar to measure the WER.
The first one (2,457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
the system, and all the rules (Constraint Grammar, lexical selection and transfer) were based on this corpus.
Hence we can call it the development corpus. It presents an upper bound on the current performance
of the system. Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN

The second corpus --- a collection of only Radio Free Europe / Radio Liberty's
Kazakh-language service articles --- was used solely for evaluation. It had a similar size --- 2,862 words.
The WER results for it can be seen in the table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.

@@

\todo{Example of lack of phon rule}

@@

To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has a near-to-production-level
coverage, but rather a prototype-level number of rules. Though the impact of this relatively low number of rules on the
quality of the translation (compare the WER results for the corpus 1 and corpus 2) looks very promising and suggests
the a high-quality translation between morphologically-rich agglutinative languages is possible.

We plan to continue development on the pair; the coverage of the system is already quite high, and although we intend to
increase it to 95\% on the corpora we have, the main work will be improving the quality of the translation by adding
more rules, starting with the Constraint Grammar module.
The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make
transfer systems between all the Turkic language pairs.  Related work is currently ongoing with Chuvash--Turkish and Turkish--Kyrgyz.

@@

The work on this Kazakh-Tatar machine translation system was partially funded by the Google Summer of Code and Google Code-In programmes.
